---
title: 'An Analysis of the ToothGrowth Data Set: Comparing the Means and Intervals
  for Significance'
author: "S Duffy"
date: "March 9, 2016"
output: word_document
---

##Overview
in 1946, E.W. Crampton was exploring the potency of vitamin C in various food sources for the Canadian government.  This was to ensure the most effective delivery of this vitamin to the armed forces during times of war.  A portion of the resulting data is stored in the R dataset 'ToothGrowth'.  This study will test the effectiveness of two sources of vitamin C, orange juice and a vitamn C pill in the form of ascorbic acid.

##Data Exploration
In this section we will explore the ToothGrowth data set from the R library. First, loading 
the data and then some summarization.  

The data set comes from research published in the paper "The Growth of the Odontoblast of the Incisor Teeth as a Criterion of Vitamin C Intake of the Guinea Pig." By E. W. Crampton as found in The Journal of Nutrition. Excellent
reading for those with insomnia.

First load the data set and take a peak at it.

```{r load}
##load the datasets library and give the data set a short alias
library(datasets)
tg <- ToothGrowth

##dimensions of the dataset
a <- dim(tg)

```

Our data set ToothGrowth has dimensions of `r a[1]` rows by `r a[2]` columns. The structure of the data 
is thus:

``` {r str}
##structure of the dataset
str(tg)
```

The beginning and end of the dataset looks like this.  

```{r headtail}
head(tg)
tail(tg)
```

Let it be noted that len is the Length of the tooth at measurement, supp is the Supplement used where VC = Vitamin C and OJ = Orange Juice, and dose is, as expected, the dose of the supplement provided in milligrams.

The following plots the data with trendlines to get a sense of what it looks like

```{r tgplot,warning=FALSE}
library(ggplot2)
tgplot <-ggplot(tg, aes(x=dose, y=len, colour=supp)) +
            geom_point(alpha=.6) +
            geom_smooth(alpha=.2, size=1) +
            ggtitle("Tooth Growth")

tgplot
```

The plot shows two distinct trendlines for each of the supplements, with orange juice at higher levels of tooth growth for lower dosages.  But is the difference significant?

##Key Assumptions
This study assumes that orange juice and vitamin C are equally efficient at enabling tooth growth in guinea pigs.  This will be the basis for hypothesis tests.  The study will be conducted by supplement type and by dosage and will assume equal variances

##Hypothesis Tests
The null hypothesis for this test will be that each type of supplement contributes an equal amount of growth
at each dosage level.  Thus, there will be three tests run, one for each dosage level (0.5, 1.0, 2.0) and for
each the null hypothesis will state that there is no difference in the average tooth growth between orange juice and vitamin C. These will be two-sided tests assuming equal variance.

####Test 1

####Dosage level 0.5, H_o: mu_a = mu_b, alpha = 0.05

The following code will subset the data and conduct the tests using a two-sided Student's t test.

``` {r test1}
tghalf <- subset(tg,dose == .5)
test1 <- t.test(subset(tghalf,supp == 'VC')[,1],subset(tghalf,supp == 'OJ')[,1],alt="two.sided",var.equal = TRUE)
pvt1 <- test1[[3]]
cit1 <- as.data.frame(test1[[4]])
```

####Test 2

####Dosage level 1.0, H_o: mu_a = mu_b, alpha = 0.05

``` {r test2}
tgone <- subset(tg,dose == 1.0)
test2 <- t.test(subset(tgone,supp == 'VC')[,1],subset(tgone,supp == 'OJ')[,1],alt="two.sided",var.equal = TRUE)
pvt2 <- test2[[3]]
cit2 <- as.data.frame(test2[[4]])
```

####Test 3

####Dosage level 2.0, H_o: mu_a = mu_b, alpha = 0.05

``` {r test3}
tgtwo <- subset(tg,dose == 2.0)
test3 <- t.test(subset(tgtwo,supp == 'VC')[,1],subset(tgtwo,supp == 'OJ')[,1], alt="two.sided",var.equal = TRUE)
pvt3 <- test3[[3]]
cit3 <- as.data.frame(test3[[4]])
```

##Results
####Test 1
In the first test, with a dosage equal to 0.5mg, the test returns a p-value
of `r pvt1` and a confidence interval of `r cit1[1,1]` to `r cit1[2,1]`.  Because the p-value is less than
our alpha of 0.05 and the hypothesized difference in means of 0 does not lie within the confidence
interval, the null hypothesis is rejected for Test 1.

####Test 2
In the second test, with a dosage equal to 1.0mg, the test returns a p-value
of `r pvt2` and a confidence interval of `r cit2[1,1]` to `r cit2[2,1]`.  Because the p-value is less than
our alpha of 0.05 and the hypothesized difference in means of 0 does not lie within the confidence
interval, the null hypothesis is rejected for Test 2.

####Test 3
In the third test, with a dosage equal to 2.0mg, the test returns a p-value
of `r pvt3` and a confidence interval of `r cit3[1,1]` to `r cit3[2,1]`.  Because the p-value is greater than
our alpha of 0.05 and the hypothesized difference in means of 0 lies within the confidence
interval, the null hypothesis is not rejected for Test 3.

##Conclusion

Considering the given data, only when the dosage of vitamin C and orange juice reaches two milligrams is 
it likely that the results of tooth growth in guinea pigs are equal to each other.  For dosages less than two milligrams (i.e., 0.5 and 1.0) there is a significant difference in the mean values that indicates the two supplements do not provide equal growth rate for guinea pig teeth.  From the test results, and given the earlier view of the data, one can assume (somewhat unscientifically) that orange juice provides better tooth growth at lower dosages. 

