---
title: An Analysis of NOAA Storm Data to Determine Impacts of Storms to Human and
  Economic Activity
author: "Stephen Duffy"
date: "January 27, 2016"
output: pdf_document
---

##Synopsis
--In a few (2-3) sentences explain what is going to be reported on.



##Simulations
--Include English explanations of the simulations you ran, with the accompanying R code. Your explanations should make clear what the R code accomplishes.

First load needed libraries and then read the data file into R
```{r readdata}
##First, load needed libraries
library(plyr)
library(dplyr)
library(reshape2)
library(ggplot2)
options(scipen=999)

##This code looks for the zipped file in your working directory and downloads it if not present
if (!file.exists("repdata-data-StormData.csv.bz2")) {
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                  destfile = "repdata-data-StormData.csv.bz2")}

##read.csv can read .bz2 files directly, this will take some time.
noaa_data <- read.csv("repdata-data-StormData.csv.bz2")

##and a quick view of the data
str(noaa_data)
```

Then perform some cleanup of the data, such as trimming leading and trailing blanks spaces,
capitalizing columns for matching, and converting values.

```{r cleandata}

```

The NOAA data has damages to crops and property separated into two columns. The first is a
numeric value and the second is an exponent, such as "K" or "M".  This analysis needs actual
numbers and so the numeric value will be multiplied by the exponent value to get an actual
value, e.g., H = 100, K = 1000, etc.  What's more, there are other characters in the exponent
column, such as numbers, plus signs, and dashes.  For any row that has a value higher than 0 in 
the numeric column, but a non-obvious exponent, I purport to multiply that by 1000, as 97-99
percent of the exponents are "K", that is 1,000.

```{r}
##This is the justification for using a 1,000 multiplier
cropexpcount <- noaa_data %>% count(CROPDMGEXP, sort = TRUE)    
propexpcount <- noaa_data %>% count(PROPDMGEXP, sort = TRUE)
cropexpcount$Pct <- round(cropexpcount[,2] / sum(cropexpcount[2:7,2]),3)
propexpcount$Pct <- round(propexpcount[,2] / sum(propexpcount[2:17,2]),3)

crop_perc <- as.data.frame(cropexpcount)
prop_perc <- as.data.frame(propexpcount)

crop_k <- as.data.frame(cropexpcount)[2,3] * 100
prop_k <- as.data.frame(propexpcount)[2,3] * 100
```




##Results


###Sample Mean versus Theoretical Mean
--Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.



###Sample Variance versus Theoretical Variance
--Include figures (output from R) with titles. Highlight the variances you are comparing. Include text that explains your understanding of the differences of the variances.


##Distribution
--Via figures and text, explain how one can tell the distribution is approximately normal.


